---
title: "Data Visualization"
author: "Matt Fleming"
date: "8/21/20"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Visualization

### 3.1 Introduction

```{r}
# we must load the tidyverse library every session
library(tidyverse)
```

### 3.2 First Steps

QUESTION: Do cars with big engines use more gas than cars with small engines?

##### 3.2.1 The mpg data frame

```{r}
mpg
```

##### 3.2.2 Creating a ggplot

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))
```

This plot defends the hypothesis that cars with larger engines will generally have worse fuel efficiency.

##### 3.2.4 Exercises

> 1. Run ggplot(data = mpg). What do you see?

There is a blank graph since we have not mapped any data points.

```{r}
ggplot(data = mpg)
```

> 2. How many rows are in mpg? How many columns? 

234 and 11

```{r}
mpg
```

> 3. What does the drv variable describe? Read the help for ?mpg to find out.

Type of drive train i.e. front wheel drive, four wheel drive, rear wheel drive

```{r}
?mpg
```

> 4. Make a scatterplot of hwy vs cyl.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = hwy, y = cyl))
```

> 5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

It does not show a correlation but instead a random series of points. Since both variables are categorical there is no way to show how often each type of car occurs.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = class, y = drv))
```

### 3.3 Aesthetic mappings

You can convey information about your data by mapping the aesthetics in your plot to the variables in your dataset. For example, you can map the colors of your points to the class variable to reveal the class of each car.

To map an aesthetic to a variable, associate the name of the aesthetic to the name of the variable inside aes(). ggplot2 will automatically assign a unique level of the aesthetic (here a unique color) to each unique value of the variable, a process known as scaling. ggplot2 will also add a legend that explains which levels correspond to which values.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

This plot maps class to the size aesthetic, which is not reccommended and prompts a warning to appear.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

This plot maps the alpha aesthetic, which changes transparency in the mapped points.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

This plot shows us why the shape aesthetic is flawed, as only six shapes can be used at once and we lose all of the suv class points.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Using quotations allows us to manually control the aesthetic and select one aesthetic for all points.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

##### 3.3.1 Exercises

> 1. What’s gone wrong with this code? Why are the points not blue?

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

Since blue is included in the aesthetic, it maps blue as a relation between a variable and a value.

> 2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

manufacturer, model, year, cyl, trans, drv, fl, class are categorical, the rest are continuous

```{r}
?mpg
```

> 3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, size = cty))
```

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = cty))
```

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))
# This chuck is supposed to have shape = cty showing us why continuous variables cannot be mapped to shape but that won't knit so I removed it.
```

> 4. What happens if you map the same variable to multiple aesthetics?

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = cyl, size = cyl))
```

> 5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, stroke = year))
```

> 6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

The aesthetic is mapped as an equation, and points that are true satisfy the equation.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

### 3.5 Facets

Adding facets splits the plot into subsets that display one group of data from the whole set. Faceted variable must be discrete.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ class, nrow = 2)
```

Facet grid is used to create more subsets of data divided by two variables.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(drv ~ cyl)
```

. can be used in place of a variable in facet grid.

##### 3.5.1 Exercises

> 1. What happens if you facet on a continuous variable?

The continuous variable is divided by each value that is represented by the data.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ cty)
```

> 2. What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?

The empty cells represent an absence of data, for example, there are no cars with rear wheel drive and 4 cylinder engines in the data. The empty cells are similar to this graph that lacks data points on 4 cylinder rear wheel drive cars.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = drv, y = cyl))
```


> 3. What plots does the following code make? What does . do?

. takes the place of the variable you replace it with, removing a division of the facet grid. Basically a facet wrap but easier to control x and y.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(drv ~ .)
```

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(. ~ cyl)
```

> 4. Take the first faceted plot in this section:

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ class, nrow = 2)
```


What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?
While the color aesthetic allows user to visualize a third variable in the overall data set, faceting allows user to use the third variable to split the data up and look at independent graphs based on the third variable. In a larger data set, 

Faceting makes the data easier to visualize when divided by more variables. While the color aesthetic allows user to visualize different categories in the same plot, faceting divides the plots and allows user to compare the trends of similar variables in separate categories. Faceting should generally be more useful in larger datasets since individual points are harder to visualize if they are overlapping or similar. The main disadvantage of faceting is in the case of smaller datasets where the data is divided and it is more difficult to visualize trends.

> 5. Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?

In the facet wrap nrow and ncol lets you pick how many rows and columns the data will be divided into. Facet grid does not have this option since the data is automatically divided based on the amount of variables.

```{r}
?facet_wrap
```

> 6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?

It is easier to visualize the data when it is neat and wider across as opposed to a straight up and down grid that you have to scroll to see.

### 3.6 Geometric Objects

Changing the geom changes the object used to map the plot. For example, geom_bar maps bar charts.

This geom_smooth maps a line.

```{r}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy))
```

Different aesthetics work for different geoms. For example, changing shape aesthetic works for a scatterplot because you can change the shape of each point, but the line doesn't show points so the shape aesthetic won't work for geom_smooth. However, changing the line type provides a similar function for smooth. This ggplot divides the lines by drive train.

```{r}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

```{r}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```


```{r}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy, color = drv))
```

This plot has both smooth and point geoms, so it maps the line and the scatterplot, allowing user to compare the raw data with a line of best fit.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + geom_smooth(mapping = aes(x = displ, y =hwy))
```

This ggplot serves the same purpose as the last, but additionally separates the data by color.

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = drv)) + geom_smooth(mapping = aes(x = displ, y = hwy, color = drv))
```

Another way to write this code and apply certain mappings to all geoms is to write the mapping in the ggplot.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + geom_point() + geom_smooth()
```

This ggplot has universally applied mappings in addition to mappings that only apply to the geom point. With this code, you can apply mappings to the layers you want while avoiding repetition.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_smooth() + geom_point(mapping = aes(color = class))
```

## Exercises 1-5

> 1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

geom_line, geom_boxplot, geom_histogram, geom_area

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_area()
```


> 2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

scatterplot and smooth line, the se = false removes standard error so there is no shaded area showing error.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + geom_point() + geom_smooth(se = FALSE)
```

> 3. What does show.legend = FALSE do? What happens if you remove it?
Why do you think I used it earlier in the chapter?

This removes the legend or key.

```{r}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy, color = drv), show.legend = FALSE)
```

> 4. What does the se argument to geom_smooth() do?

It displays the standard margin of error.

> 5. Will these two graphs look different? Why/why not?

They will look the same because they both have the same mappings, but the first universally applies the mappings while the second individually applies them.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point() + geom_smooth()
```

```{r}
ggplot() + geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

### 3.7 Statistical Transformations

```{r}
?diamonds
```

```{r}
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut))
```

### 3.8 Position Adjustments

```{r}
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, fill = cut))
```

```{r}
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, fill = clarity))
```

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + geom_bar(alpha = 0.2, position = "identity")
```

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, color = clarity)) + geom_bar(fill = NA, position = "identity")
```

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + geom_bar(position = "fill")
```

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + geom_bar(position = "dodge")
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point()
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point(position = "jitter")
```

##### 3.8.1 Exercises

> 1. What is the problem with this plot? How could you improve it?

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_point()
```




